---
title: "Correspondence analyses of the Mueller Report"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r packages, echo=FALSE, warning=FALSE, message=FALSE}
library(ExPosition)
  # if needed.
# devtools::install_github("derekbeaton/GSVD")
# devtools::install_github("derekbeaton/OuRS", subdir = "/OuRS")
library(GSVD)
library(ours)

library(ggplot2)
library(gridExtra)
library(ggrepel)
library(corrplot)
```

# The gist

* The Mueller Report recently dropped and #rstats community quickly went to work.
* I saw two analyses on word frequencies ([by Mara Averick](https://github.com/batpigandme/tidymueller)) and co-occurence networks ([by Chris Bail](https://github.com/cbail/mueller_report)).
* I wanted to apply correspondence analyses to slightly modified versions of their data
  * akin to how [we analyzed the NeuroSynth database](https://www.biorxiv.org/content/10.1101/157826v3).
  * Correspondence analysis is a technique like principal components analysis, but designed for categorical, count, non-negative, and similar (i.e., χ2-friendly) data
* The goals of the analyses were several fold:
  * Treat the data in a bi-variate way to get information for *both* the pages and the words/lemmas.
  * Assess the similarity and grouping of *pages* within each data set, i.e., with respect to preprocessing/preparation of the data
  * Compute the similiarity between the component structures from each analysis, i.e., do we see similar or different component structures based on the different versions (words vs. lemmas) of the data.

## Short version of the results

* Surprisingly: Both versions of the data produce effectivley identical results for approximately the first 8 (possibly more) components
  * I did not expect this because the two data sets are different (worsd are individual words, lemmas are particular words, letters/keys, or phrases/n-grams)
* The beginning and end pages of the Mueller Report are effectively orthogonal to one another, with highly specific words/lemmas for each
* There is an extraordinary density of repeated terms in the middle of the report

# Methods

## Cleaning and preprocessing

In `./R/0_read_preprocess.save.R`, I use the same base cleaning steps as in [batpigandme/tidymueller](https://github.com/batpigandme/tidymueller) and [cbail/mueller_report](https://github.com/cbail/mueller_report). I use some slight modifications of those for extra cleaning and maintaining particular data features or the data shapes. This file exports some `.rda` files for subsequent use.

In `./R/1_make_clean_matrices.R`, I reshape the data into a rectangle (pages x words or pages x lemmas), then prune out any excessively infrequent rows and columns, and finally export the "analysis ready" matrices as `.csv` files. 

## Analyses

All analyses were performed here in `README.rmd`. I applied correspondence analysis (CA) to two versions of the data:
* pages x words: each cell contains the frequency of that word on that page (via [batpigandme/tidymueller](https://github.com/batpigandme/tidymueller))
* pages x lemmas: each cell contains the frequency of that lemma on that page (via [cbail/mueller_report](https://github.com/cbail/mueller_report))

I used CA from one of my packages: [ExPosition](https://cran.r-project.org/web/packages/ExPosition/index.html). CA is a technique like principal components analysis (PCA), but designed for categorical, count, non-negative, and similar (i.e., χ2-friendly) data. When data are proportional, compositional, and generally χ2-friendly, PCA is not suitable. That's because PCA wants to compute a covariance (or correlation) matrix from the data. In this case, the data should not (cannot!) be normalized down the columns. CA provides a χ2-projection onto a Euclidean space. CA also has innumerable nearly magical properties (one of those magical properties is that if you have a network with a clique, CA produces eigenvalues of exactly 1 for every clique; that's something for later...).

Like PCA, CA produces orthogonal components that we rank order by their explained variance. So, the first component explains the most amount of variance, while the second component explains the next most amount of possible variance, where the second component is orthogonal to the first. The third component explains the next most variance after the second, where the third component is orthogonal to the second (and thus the first), and so on. For each of these components we can plot the component scores for the rows (pages) and columns (words or lemmas) separately on component maps. Because CA is a bi-variate technique, so the scales of the component scores for the rows and columns are the same (with caveats; also something for later!).

These analyses were done for two reasons: (1) emphasize the pages (not the words), and (2) assess how similar the results are from each type of text preprocessing. Other analyses have focused on words and word frequencies. While that is part of the results here, it's not much different from what other analyses find. Instead, the emphasis here is on the pages. We're using CA here to extract the components of pages and words, but we can take a closer look at the pages to see if some kind of structure emerges from the similarity of pages. Once we've done both analyses, we can then find out if processing the document in two relatively different ways produces similar or different results, with respect to the overall structure of page similarity. 

Results were plotted with [ggplot2](https://ggplot2.tidyverse.org/) and [corrplot](https://cran.r-project.org/web/packages/corrplot/index.html). Scree plot code adapted (mostly stolen!) from [Matt Kmiecik](https://mattkmiecik.com/post-Exploring-11-Years-of-Chicago-Blackhawks-Data-using-Principal-Components-Analysis.html). 

# Correspondence analyses of the Mueller Report

```{r load_data_run_cas, echo=FALSE, warning=FALSE, message=FALSE}

page_by_word_counts.matrix_for_analyses <- read.csv(here::here("data","for_analyses","page_by_word_counts.matrix_for_analyses.csv"), row.names = 1)
page_by_word.matrix_for_analyses <- read.csv(here::here("data","for_analyses","page_by_word.matrix_for_analyses.csv"), row.names = 1)

page_by_lemma_counts.matrix_for_analyses <- read.csv(here::here("data","for_analyses","page_by_lemma_counts.matrix_for_analyses.csv"), row.names = 1)
page_by_lemma.matrix_for_analyses <- read.csv(here::here("data","for_analyses","page_by_lemma.matrix_for_analyses.csv"), row.names = 1)

words_counts_ca <- epCA(page_by_word_counts.matrix_for_analyses, graphs=F)
# words_ca <- epCA(page_by_word.matrix_for_analyses, graphs=F)
# words_gen.mcd <- cat.mcd(t(page_by_word.matrix_for_analyses), make.data.disjunctive = F, alpha = .75)

lemmas_counts_ca <- epCA(page_by_lemma_counts.matrix_for_analyses, graphs=F)
# lemmas_ca <- epCA(page_by_lemma.matrix_for_analyses, graphs=F)
# lemmas_gen.mcd <- cat.mcd(t(lemmas_ca), make.data.disjunctive = F, alpha = .75)



```

The pages x words data matrix was `r nrow(page_by_word_counts.matrix_for_analyses)` rows (pages) by `r ncol(page_by_word_counts.matrix_for_analyses)` columns (words). The pages x lemmas data matrix was `r nrow(page_by_lemma_counts.matrix_for_analyses)` rows (pages) by `r ncol(page_by_lemma_counts.matrix_for_analyses)` columns (lemmas).

## Explained variance

```{r summaries_diagnoistics, echo=FALSE, warning=FALSE, message=FALSE}

words_counts_ca_scree <- data.frame(eigs = words_counts_ca$ExPosition.Data$eigs,
                perc_explained = words_counts_ca$ExPosition.Data$t,
                comps = 1:length(words_counts_ca$ExPosition.Data$eigs)
                )

# words_ca_scree <- data.frame(eigs = words_ca$ExPosition.Data$eigs,
#                 perc_explained = words_ca$ExPosition.Data$t,
#                 comps = 1:length(words_ca$ExPosition.Data$eigs)
#                 )

lemmas_counts_ca_scree <- data.frame(eigs = lemmas_counts_ca$ExPosition.Data$eigs,
                perc_explained = lemmas_counts_ca$ExPosition.Data$t,
                comps = 1:length(lemmas_counts_ca$ExPosition.Data$eigs)
                )

# lemmas_ca_scree <- data.frame(eigs = lemmas_ca$ExPosition.Data$eigs,
#                 perc_explained = lemmas_ca$ExPosition.Data$t,
#                 comps = 1:length(lemmas_ca$ExPosition.Data$eigs)
#                 )


words_counts_ca_scree_plot <- ggplot(words_counts_ca_scree, aes(factor(comps), perc_explained)) +
  geom_point() +
  geom_path(aes(group = 1)) +
  labs(x = "Components", y = "Explained Variance (%)") +
  ggtitle("Scree Plot: Pages x Words (Counts)") +
  theme(axis.ticks.x = element_blank(), axis.text.x = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background = element_blank())

# words_ca_scree_plot <- ggplot(words_ca_scree, aes(factor(comps), perc_explained)) +
#   geom_point() +
#   geom_path(aes(group = 1)) +
#   labs(x = "Components", y = "Explained Variance (%)") +
#   ggtitle("Scree Plot: Pages x Words (Binary)") +
#   theme(axis.ticks.x = element_blank(), axis.text.x = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background = element_blank())
  
lemmas_counts_ca_scree_plot <- ggplot(lemmas_counts_ca_scree, aes(factor(comps), perc_explained)) +
  geom_point() +
  geom_path(aes(group = 1)) +
  labs(x = "Components", y = "Explained Variance (%)") +
  ggtitle("Scree Plot: Pages x Lemmas (Counts)") +
  theme(axis.ticks.x = element_blank(), axis.text.x = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background = element_blank())

# lemmas_ca_scree_plot <- ggplot(lemmas_ca_scree, aes(factor(comps), perc_explained)) +
#   geom_point() +
#   geom_path(aes(group = 1)) +
#   labs(x = "Components", y = "Explained Variance (%)") +
#   ggtitle("Scree Plot: Pages x Lemmas (Binary)") +
#   theme(axis.ticks.x = element_blank(), axis.text.x = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background = element_blank())

# grid.arrange(words_counts_ca_scree_plot, words_ca_scree_plot, lemmas_counts_ca_scree_plot, lemmas_ca_scree_plot, nrow = 2)

```

Each analysis produces more than 400 components. The explained variance per component can be visualized with a Scree plot, which shows the components in descending (explained variance) order from the first to the last. The Scree plot tells us how much variance each component explains, and where the "Scree" (junk at the bottom of a slope) happens. 

```{r vis_scree,echo=FALSE, warning=FALSE, message=FALSE, fig.height=6, fig.width=12}
grid.arrange(words_counts_ca_scree_plot, lemmas_counts_ca_scree_plot, nrow = 1)
```

Both analyses show that the first 15 or so components explain considerably more variance (proportionally) than the remaining components. The "Scree" part happens at around 100 components (the bend) but I'm not going to interpret that many components (this is an exercise for the reader).

## Component maps

```{r component_vis_prep, echo=FALSE, warning=FALSE, message=FALSE}

words_counts_ca_comps12_fi_df <- data.frame(
  COMP1 = words_counts_ca$ExPosition.Data$fi[,1],
  COMP2 = words_counts_ca$ExPosition.Data$fi[,2],
  PAGE_NUMBER = as.numeric(rownames(words_counts_ca$ExPosition.Data$fi))
)
words_counts_ca_comps12_fj_df <- data.frame(  
  COMP1 = words_counts_ca$ExPosition.Data$fj[,1],
  COMP2 = words_counts_ca$ExPosition.Data$fj[,2]
)
words_counts_xlims <- c(
  min(c(0.01,words_counts_ca_comps12_fi_df$COMP1,words_counts_ca_comps12_fj_df$COMP1)),
  max(c(-0.01,words_counts_ca_comps12_fi_df$COMP1,words_counts_ca_comps12_fj_df$COMP1))
)
words_counts_ylims <- c(
  min(c(0.01,words_counts_ca_comps12_fi_df$COMP1,words_counts_ca_comps12_fj_df$COMP2)),
  max(c(-0.01,words_counts_ca_comps12_fi_df$COMP1,words_counts_ca_comps12_fj_df$COMP2))
)

lemmas_counts_ca_comps12_fi_df <- data.frame(
  COMP1 = lemmas_counts_ca$ExPosition.Data$fi[,1],
  COMP2 = lemmas_counts_ca$ExPosition.Data$fi[,2],
  PAGE_NUMBER = as.numeric(rownames(lemmas_counts_ca$ExPosition.Data$fi))
)
lemmas_counts_ca_comps12_fj_df <- data.frame(
  COMP1 = lemmas_counts_ca$ExPosition.Data$fj[,1],
  COMP2 = lemmas_counts_ca$ExPosition.Data$fj[,2]
)
lemmas_counts_xlims <- c(
  min(c(0.01,lemmas_counts_ca_comps12_fi_df$COMP1,lemmas_counts_ca_comps12_fj_df$COMP1)),
  max(c(-0.01,lemmas_counts_ca_comps12_fi_df$COMP1,lemmas_counts_ca_comps12_fj_df$COMP1))
)
lemmas_counts_ylims <- c(
  min(c(0.01,words_counts_ca_comps12_fi_df$COMP1,lemmas_counts_ca_comps12_fj_df$COMP2)),
  max(c(-0.01,words_counts_ca_comps12_fi_df$COMP1,lemmas_counts_ca_comps12_fj_df$COMP2))
)



words_counts.pages_component_map <- 
  ggplot(words_counts_ca_comps12_fi_df, aes(COMP1, COMP2, color=PAGE_NUMBER)) +
  geom_hline(yintercept = 0, color = "black") +
  geom_vline(xintercept = 0, color="black") +
  geom_point() +
  scale_color_gradient(low="blue", high="red") +
  xlim(words_counts_xlims[1] * 1.1, words_counts_xlims[2] * 1.1) +
  ylim(words_counts_ylims[1] * 1.1, words_counts_ylims[2] * 1.1) +
  xlab(paste0("Component 1. Explained Variance: ", round(words_counts_ca$ExPosition.Data$t[1],digits=2), "%" ) ) +
  ylab(paste0("Component 2. Explained Variance: ", round(words_counts_ca$ExPosition.Data$t[2],digits=2), "%" ) ) +
  ggtitle("Pages x Words. Component Map of Pages.") +
  theme(axis.line = element_blank(),
    panel.border = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks = element_blank(), 
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank(), 
    panel.background = element_blank())


words_counts.words_component_map <- ggplot(words_counts_ca_comps12_fj_df, aes(COMP1, COMP2, label=rownames(words_counts_ca_comps12_fj_df))) +
  geom_hline(yintercept = 0, color = "black") +
  geom_vline(xintercept = 0, color="black") +
  geom_point(color="black",fill="olivedrab3", alpha=.7, shape=21) +
  xlim(words_counts_xlims[1] * 1.1, words_counts_xlims[2] * 1.1) +
  ylim(words_counts_ylims[1] * 1.1, words_counts_ylims[2] * 1.1) +
  xlab(paste0("Component 1. Explained Variance: ", round(words_counts_ca$ExPosition.Data$t[1],digits=2), "%" ) ) +
  ylab(paste0("Component 2. Explained Variance: ", round(words_counts_ca$ExPosition.Data$t[2],digits=2), "%" ) ) +
  ggtitle("Pages x Words. Component Map of Words.") +
  theme(axis.line = element_blank(),
    panel.border = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks = element_blank(), 
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank(), 
    panel.background = element_blank())


lemmas_counts.pages_component_map <- ggplot(lemmas_counts_ca_comps12_fi_df, aes(COMP1, COMP2, color=PAGE_NUMBER)) +
  geom_hline(yintercept = 0, color = "black") +
  geom_vline(xintercept = 0, color="black") +
  geom_point() +
  scale_color_gradient(low="blue", high="red") +
  xlim(lemmas_counts_xlims[1] * 1.1, lemmas_counts_xlims[2] * 1.1) +
  ylim(lemmas_counts_ylims[1] * 1.1, lemmas_counts_ylims[2] * 1.1) +
  xlab(paste0("Component 1. Explained Variance: ", round(lemmas_counts_ca$ExPosition.Data$t[1],digits=2), "%" ) ) +
  ylab(paste0("Component 2. Explained Variance: ", round(lemmas_counts_ca$ExPosition.Data$t[2],digits=2), "%" ) ) +
  ggtitle("Pages x Lemmas. Component Map of Pages") +
  theme(axis.line = element_blank(),
    panel.border = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks = element_blank(), 
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank(), 
    panel.background = element_blank())


lemmas_counts.lemmas_component_map <- ggplot(lemmas_counts_ca_comps12_fj_df, aes(COMP1, COMP2, label=rownames(lemmas_counts_ca_comps12_fj_df))) +
  geom_hline(yintercept = 0, color = "black") +
  geom_vline(xintercept = 0, color="black") +
  geom_point(color="black",fill="olivedrab3", alpha=.7, shape=21) + 
  xlim(lemmas_counts_xlims[1] * 1.1, lemmas_counts_xlims[2] * 1.1) +
  ylim(lemmas_counts_ylims[1] * 1.1, lemmas_counts_ylims[2] * 1.1) +
  xlab(paste0("Component 1. Explained Variance: ", round(lemmas_counts_ca$ExPosition.Data$t[1],digits=2), "%" ) ) +
  ylab(paste0("Component 2. Explained Variance: ", round(lemmas_counts_ca$ExPosition.Data$t[2],digits=2), "%" ) ) +
  ggtitle("Pages x Lemmas Component Map of Lemmas.") +
  theme(axis.line = element_blank(),
    panel.border = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks = element_blank(), 
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank(), 
    panel.background = element_blank())


```

When we plot pages (or words, or lemmas) on Component maps, we can identify overall structures within the data. We can also see how various pages (or words, or lemmas) are related. On the components, if we see any particular pages, words, or lemmas (a.k.a "items") near one another, then we know they are positively correlated. If we see any items far apart from one another along a line, then they are negatively correlated. If we see any items near or at a 90 degree angle, then they are orthogonal to one another.

```{r vis_components, echo=FALSE, warning=FALSE, message=FALSE, fig.width=12, fig.height=12}
grid.arrange(words_counts.pages_component_map, words_counts.words_component_map,
             lemmas_counts.pages_component_map, lemmas_counts.lemmas_component_map,
             nrow=2)
```

The above component maps show the pages x words analysis on the top and the pages x lemmas on the bottom. On the left side we see the component scores for the pages and on the right we see the component scores for the words (top) and lemmas (bottom). The pages have been colored along a gradient. We're only going to focus on the pages (the words are more numerous and dense). 

First, I'm not into component/factor rotations, but these components scream "rotate me!". Interestingly, we see a nearly perfect orthogonal structure emerge for both, but just slightly off from the components (axes). Second, both analyses show quite an interesting structure of the pages: the beginning of the document (blue) is at a near perfect right angle to the end of the document (red). Furthermore, the beginnings and ends of the document are unique, as we can see that these are very far from the center of the graph (origin). Thus the words found on those pages are both (1) rarely used and (2) exist almost exclusively on those pages. As for the middle (purple-ish dots) of the document, there are frequent uses of many of the same words. 

## Similarity between analyses

Given that the two analyses were on two differently processed data sets, I was curious if these two data sets produced similar or different results. To do so, I computed the correlations (and visualized with a correlogram via `corrplot`) between the component scores of each analysis for the pages common to both. Some pages for each data set were dropped because of excessively low frequency of words or lemmas.

```{r vis_corrplot, echo=FALSE, warning=FALSE, message=FALSE, fig.width=8, fig.height=8}

common.pages <- intersect(rownames(lemmas_counts_ca$ExPosition.Data$fi),rownames(words_counts_ca$ExPosition.Data$fi))
corrplot(cor(words_counts_ca$ExPosition.Data$fi[common.pages,],lemmas_counts_ca$ExPosition.Data$fi[common.pages, ])[1:20,1:20]^2,cl.lim=c(0,1))



```

The `corrplot` above shows that the first 8 (possibly 11, or 13) components are virtually identical. I did not expect that! It seems that for both of these preprocessing approaches, the low rank same structure emerges. Between the first 8 components, the Rv-coefficient (akin to a multivariate R-squared) was `r  round(rvCoeff(cor(words_counts_ca$ExPosition.Data$fi[common.pages,1:8]),cor(lemmas_counts_ca$ExPosition.Data$fi[common.pages,1:8])), digits=3)`.

# Conclusions & More Analyses

So what can we conclude from this? First is that both preprocessing approaches produce almost identical results by the end of analyses. This suggests that the granularity of the words is sufficient enough compared to the lemmas. But that also means that the separate words in phrases picked up by the lemmas are also likely co-occuring in high frequency both within and across pages.

We also see that, generally, there are completely different word usages at the beginning and end of the document. Other components may also help tease apart where highly similar (or nearly orthogonal) sections of the document exist (again: an exercise for the reader).

Also just amongst these analyses are a lot of results to interpret. For example, what are the words/lemmas that underly the page structure that makes the beginning and end so orthogonal? (another exercise for the reader!)

As for more analyses, I had hinted as some *very* cool properties of CA when it comes to graphs. In the future (the distant future, like 1999 or so) I plan to come back to these data because this is a highly rich dataset that can be analyzed in a number of ways. There are two particular branches of analyses for the future:

* A thresholded version of these data (0 vs. 1 for just absence vs. presence of a word; no need for counts) is effectively a graph. CA has nearly magical properties that allows us to detect global network structures and cliques (if they exist) from either the crossproduct (adjacency or weighted adjancency) matrices or just directly on the rectangular (binary) matrices (both ways produce identical results). To note: the current results are also effectively showing what amounts to bi-variate graph structures: a global network structure for the pages and a global network structure for the words/lemmas.

* Analyses to detect anomalous pages and words. With a technique we [recently developed](https://www.biorxiv.org/content/10.1101/333005v2.abstract) (that is an extension of the minimum covariance determinant) we can identify both (1) the robust sub-structure in the data and (2) any anomalous (outlier) words.


